---
title: "DS325 Project 3: Recidivism"
author: "Johnny Carstens & Matthew Bushnell"
date: "12/2/2021"
output:
  html_document:
    theme: spacelab
    highlight: haddock
    toc: true
    toc_float: true
    number_sections: false
    code_folding: hide
---

## Introduction

## Logistic Regression of Recidivism Data
# Task 1: Visualizing the Data


```{r trapAnthem}
library(tidyverse)
library(olsrr)
library(ggfortify)
library(GGally)
library(mdsr)

ggplot(pairs, aes(x = age, fill = as.factor(isRecid), alpha = 0.1)) + 
  geom_density() + 
  labs(x = "Age", y = "isRecid", title = "Figure 1: Age vs isRecid")
```

Figure 1 is a density plot of Age vs isRecid. isRecid is a binary response variable where 1 is equal to a person who did reoffend, and 0 is equal to someone who hasn't reoffended. Notice that in Figure 1, as age increases the amount of people who reoffended, or didn't reoffend decreases. We can also see that between ages 30 and 35, the amount of people who didn't recidivate becomes greater than those who did recidivate. 

```{r jPop}
ggplot(pairs, aes(x=priorsCount, fill=as.factor(isRecid), alpha = 0.1)) + 
  geom_density() + 
  labs(x="Number of Prior Convictions", y = "isRecid", title = "Figure 2: isRecid vs Number of Prior Convictions")
```

Figure 2 is a density plot of the number of prior convictions who did and did not recidivate. This data is fairly skewed, so we apply the natural log function to the number of prior convictions which brings us to the next plot. 

```{r Dragostea Din Tei}
ggplot(pairs, aes(x=log(priorsCount), fill=as.factor(isRecid), alpha = 0.1)) + 
  geom_density() + 
  labs(x="Number of Prior Convictions", y = "isRecid", title = "Figure 2: isRecid vs Number of Prior Convictions")
```

```{r I cant remember if we wanted to include this one}
ggplot(pairs, aes(x=sex, fill = as.factor(isRecid))) + geom_bar(position = "dodge") + 
  labs(x = "Sex", y = "Count", title = "Figure 3: isRecid vs Sex")
```

