---
title: "DS325 Project 3: Recidivism"
author: "Johnny Carstens & Matthew Bushnell"
date: "12/2/2021"
output:
  html_document:
    theme: spacelab
    highlight: haddock
    toc: true
    toc_float: true
    number_sections: false
    code_folding: hide
---

## Introduction


## Task 1: Visualize the Data

```{r trapAnthem}
library(tidyverse)
library(olsrr)
library(ggfortify)
library(GGally)
library(mdsr)

ggplot(pairs, aes(x=RiskRecidDecileScore, fill = as.factor(isRecid))) + geom_bar(position = "dodge") + 
  labs(x = "RiskRecidDecileScore", y = "Count", title = "Figure 1: RiskRecidDecileScore vs isRecid")
```
Figure 1 is a bar graph that counts how many people either did or did not recidivate (Red means they did not recidivate) for each score of the COMPAS test. This plot shows a clear correlation that people with a low score did not recidivate. There isn't a very clear relation for people who did recidivate with respect to people that did recidivate. The plot shows that RiskRecidDecileScore is a good indicator of prediciting if a person will not recidivate, so we will include this in our model. 

```{r jPop}
ggplot(pairs, aes(x=priorsCount, y = isRecid, color = race, alpha = 0.1)) + geom_point() + geom_jitter(aes(height = 0.5)) 
```


```{r I cant remember if we wanted to include this one}
ggplot(pairs, aes(x=sex, fill = as.factor(isRecid))) + geom_bar(position = "dodge") + 
  labs(x = "Sex", y = "Count", title = "Figure 3: isRecid vs Sex")
```
